---
title: "project_tutorial"
output: 
  pdf_document:
    latex_engine: xelatex
    keep_tex: true
vignette: >
  %\VignetteIndexEntry{project_tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


Hello. A lot of development is currently underway.
As of 4/11/2025 The following needs to get done by 4/14/2025:

1. Re-Writing/Re-Directing RNA-Seq Pre-Processing Pipeline
2. Running Real Rna-Seq data-submitted on 4/11/2025
3. Running the Join Script on Real Data
4. Testing MEME
3. Docking each of these scripts
Until then this will be majority text based guide


## SnakeMake Pre-Processing Commands
 

UNIX commands (Ran on the Umich Great Lakes Cluster)

Before running
Ensure your directory tree looks similar to the example below

snakemake/
|── Snakefile                      # Main workflow
|── config/
│   └── config.yaml                # Parameters, sample info, etc.
|─ envs/
│   └── chip_pipeline.yaml        # Conda environments
|── scripts/
│   ├── custom_analysis.R         # R analysis scripts
│   └── helper_functions.py       # Python helpers
|── data/
│   └── chip_data/
|                           # Raw fastq files, e.g., SRRxxxx_1.fastq
|── chip_pipleine/
    └── SnakeFile/            #SnakeFile to be called
    
Repeat this structure for RNA-Seq Pre-Processing Pipeline


You should alread have conda installed, else: https://docs.conda.io/projects/conda/en/stable/user-guide/install/linux.html
Install SnakeMake if you have not already
# Linux
wget https://repo.anaconda.com/miniconda/Miniconda3-latest-Linux-x86_64.sh
bash Miniconda3-latest-Linux-x86_64.sh

# macOS
wget https://repo.anaconda.com/miniconda/Miniconda3-latest-MacOSX-x86_64.sh
bash Miniconda3-latest-MacOSX-x86_64.sh

Creat a Cona Env
conda create -n snakemake-env -c conda-forge -c bioconda snakemake
conda activate snakemake-env


Make your environments
conda config --set channel_priority strict
or that does not work channel_priority flexible


cd to the env folder
conda env create --name rna_pipeline --file rna_pipeline.yaml
conda env create --name chip_pipeline --file chip_pipeline.yaml

Make Sure you have the paired-end Sequences (ChIP/CUT&TAG, RNA-seq)
into their seperate folders under ./data/chip_data, and ./data/rna_data

conda activate one of the environments

conda activate rna_pipeline or conda activate rna_pipeline
### SnakeMake Commands

cd to the respective pre-processing folder

run using snakemake --dry-run to check the pipeline is working

snakemake --cores 32 # for the real datasets

As of 4/11/2025 we must use some other datasets until the real ones are ready

We are using a real dataset but not the one are planning.

```{r}
library(postprocess)
library(DESeq2)
library(tidyverse)
library(clusterProfiler)
library(org.Hs.eg.db)
library(biomaRt)
```


Load the referance for enrichr
```{r}
#this function needs to be generalized. How? Need to take a page out of the clusterprofiler
mmu_entrezid_to_genename <- function(entrez_vector) {
  #requires both of these packages to run
  #require("biomaRt")
  #require("org.Mmu.eg.db")
  #connect to the Ensembl database for Mmu
  ensembl = useMart("ensembl", dataset = "mmusculus_gene_ensembl")
  #getting the corresponding Entrez Gene IDs
  entrezid_genename_df <- getBM(attributes = c("entrezgene_id", "external_gene_name"), filters = "entrezgene_id", values = entrez_vector, mart = ensembl)
  #return both the df and the vector
  return(entrezid_genename_df)
}
```


Conversion Helper Function
```{r}

# common example datasets: "hsapiens_gene_ensembl", "mmusculus_gene_ensembl"
#connects to the ensembledb, grabs and returns
#arguments: an entrez id vector, and the name of ensembldataset
#It would be wise to load in your chosen OrgDb set before running this code

entrezid_to_genename <-function(entrez_vector,
                                ensembl_dataset = "hsapiens_gene_ensembl"){
                          #connect to the Ensembl database for Mmu
                          ensembl = useMart("ensembl", dataset = ensembl_dataset)
                          #getting the corresponding Entrez Gene IDs
                          entrezid_genename_df <- getBM(attributes = c("entrezgene_id", "external_gene_name"), filters = "entrezgene_id", values = entrez_vector, mart = ensembl)
    return(entrezid_genename_df)
}
```
Using example data
Input should be unormalized data counts

Data Source : https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE60424

Importing Meta Data from data

```{r}
conditions <- c(
  "Healthy Control", "MS pretreatment", "MS posttreatment", "Type 1 Diabetes", "Type 1 Diabetes",
  "Healthy Control", "Healthy Control", "Healthy Control", "Healthy Control", "Healthy Control",
  "Healthy Control", "MS pretreatment", "MS pretreatment", "MS pretreatment", "MS pretreatment",
  "MS pretreatment", "MS posttreatment", "MS posttreatment", "MS posttreatment", "MS posttreatment",
  "MS posttreatment", "Type 1 Diabetes", "Type 1 Diabetes", "Type 1 Diabetes", "Type 1 Diabetes",
  "Type 1 Diabetes", "Type 1 Diabetes", "Type 1 Diabetes", "Type 1 Diabetes", "Type 1 Diabetes",
  "Type 1 Diabetes", "Type 1 Diabetes", "Sepsis", "Sepsis", "Sepsis", "Sepsis", "Sepsis", "Sepsis",
  "Sepsis", "Sepsis", "Sepsis", "Sepsis", "Sepsis", "Sepsis", "Sepsis", "Sepsis", "Sepsis", "Sepsis",
  "Sepsis", "Sepsis", "Sepsis", "Sepsis", "MS pretreatment", "MS pretreatment", "MS pretreatment",
  "MS pretreatment", "MS pretreatment", "MS pretreatment", "MS pretreatment", "MS posttreatment",
  "MS posttreatment", "MS posttreatment", "MS posttreatment", "MS posttreatment", "MS posttreatment",
  "MS posttreatment", "Healthy Control", "Healthy Control", "Healthy Control", "Healthy Control",
  "Healthy Control", "Healthy Control", "Healthy Control", "Healthy Control", "Healthy Control",
  "Healthy Control", "Healthy Control", "Healthy Control", "Healthy Control", "Healthy Control",
  "ALS", "ALS", "ALS", "ALS", "ALS", "ALS", "ALS", "Healthy Control", "Healthy Control",
  "Healthy Control", "Healthy Control", "Healthy Control", "Healthy Control", "Healthy Control",
  "ALS", "ALS", "ALS", "ALS", "ALS", "ALS", "ALS", "Type 1 Diabetes", "Type 1 Diabetes",
  "Type 1 Diabetes", "Type 1 Diabetes", "Type 1 Diabetes", "Type 1 Diabetes", "Type 1 Diabetes",
  "Type 1 Diabetes", "Type 1 Diabetes", "Type 1 Diabetes", "Type 1 Diabetes", "Type 1 Diabetes",
  "Type 1 Diabetes", "MS pretreatment", "MS pretreatment", "MS pretreatment", "MS pretreatment",
  "MS pretreatment", "MS pretreatment", "MS posttreatment", "MS posttreatment", "MS posttreatment",
  "MS posttreatment", "MS posttreatment", "MS posttreatment", "MS posttreatment", "ALS", "ALS",
  "ALS", "ALS", "ALS", "ALS", "ALS"
)
```

Importing Raw Counts Data

User must assemble the following pieces of data before using the functions in this package:

1: They must have pre-processed RNa-Seq in a raw un-normalized count matrix and ChIP-Seq data in the Annotated Peak Call Form

2: They must have meta data which is equal in length to the sample name vector (Use will probably have to search.


```{r}
#load in r
toy_data <- read_delim("../../toy_data/GSE60424_raw_counts_GRCh38.p13_NCBI.tsv", delim = "\t")
toy_data <- as.data.frame(toy_data)
#Set GeneID as row names 
counts <- column_to_rownames(toy_data, var = "GeneID")

#Create metadata 
metadata <- data.frame(condition = conditions, row.names = colnames(counts))
```

Should be converted to an Rscript from here on out

Running Deseq2
```{r}
dds <- DESeqDataSetFromMatrix(
  countData = counts,
  colData = metadata,
  design = ~ condition
)
dds <- DESeq(dds)

normalized_counts <- counts(dds, normalized = TRUE)

results_table <- results(dds)

results_frame <- as.data.frame(results_table)
```

User inputs: countsData = user read in data, colData = metadata, design = conditions of each sample included.

Note: Inputs: countData = user read in data


```{r}
results_frame <- as.data.frame(results_table)
results_frame
```

Zeroth Function Chip_seq and RNA_seq Merger (Requirements: same experiment/organism and annotations)

First Function: postprocess::deseq_filtration

step 1, applying pvalue arguments

provisional function data to txt


```{r}
deseq_filteration <- function(tibble, 
                              file_name, 
                              pval_cutoff, 
                              rev = FALSE){
#added argument: pval_cutoff, allows user to set their pvalue cutoff
# This function takes into consideration that the Deseq2 Results where flipped
# I'll have to make a conditional on that front
#
library(tidyverse)
  
  if(!require(openxlsx)) install.packages("openxlsx", dependencies=TRUE)
  
  if(!require(dplyr)) install.packages("dplyr", dependencies=TRUE)
  
library(dplyr)
library(openxlsx)
  
wbfxn <- createWorkbook()

  addWorksheet(wbfxn, paste("Initial Data", nrow(tibble)))
  writeData(wbfxn, sheet = paste("Initial Data", nrow(tibble)) , tibble)

tibble_exp <- tibble %>%
# filter in value <=0.05
  
#In the past I have encountered Deseq sets wherelog2FCs where the results are flipped
#This conditional takes that into consideration
  {
    if (rev) {
      . <- mutate(., INUN_log2FC = log2FoldChange * -1)
      . <- relocate(., INUN_log2FC, .after = 2)
      addWorksheet(wbfxn, paste("Corrected Log2FC", nrow(.)))
      writeData(wbfxn, sheet = paste("Corrected Log2FC", nrow(.)), .)
    } else {
      . <- mutate(., log2FoldChange = log2FoldChange)
    }
    
  .}%>%
  filter(pvalue < pval_cutoff) %>%
  {
    addWorksheet(wbfxn, paste("pvalue <=", pval_cutoff, nrow(.)))
    writeData(wbfxn, sheet = paste("pvalue <=", pval_cutoff, nrow(.)), .)
    
  .}%>%
  
  {
    if ("INUN_log2FC" %in% colnames(.)) {
      mutate(., plusFC = 2^INUN_log2FC)
    } else {
      mutate(., plusFC = 2^log2FoldChange)
    }
  } %>%
  
  relocate(plusFC, .before = 4) %>%
  {
    
addWorksheet(wbfxn, "add plusFC") 
writeData(wbfxn,sheet = "add plusFC", .) 

  .}%>%
  
  arrange(desc(plusFC)) %>%
  {
addWorksheet(wbfxn, "Sort by plusFC desc.") 
writeData(wbfxn, sheet = "Sort by plusFC desc.", .)

  .}%>%
  
#add a column called negFC: if a value in plusFC is less than 0.05 then convert it to
#negFC (-1/FC) else leave the value alone by multiplying it by 1
  mutate(negFC = if_else(plusFC < 0.5, -1/plusFC, plusFC * 1)) %>%
  relocate(negFC, .before = 5) %>%
  {
addWorksheet(wbfxn, "add negFC") 
writeData(wbfxn, sheet = "add negFC", .)
  .}%>%
  
#In a new column called FC, 
#If the values in plusFC and negativeFC match (plusFC == negFC) then copy the values that do match into the new column and convert them to the character type 
#Else: paste both values from negFC and plusFC into the same point and seperate them by a "/" character.
  mutate(FC = if_else(plusFC == negFC, as.character(plusFC), paste(plusFC, negFC, sep = "/"))) %>%
  #In the new column FinalFC, If an entry has a slash charcter in the column FC, then get rid of everycharacter before the slash in each entry including the slash.
  mutate(FinalFC = gsub(".*?/","", FC)) %>%
  #convert the column FinalFC into a numeric type variable
  #The goal of the FinalFC column is only keep the strongest directional fold change
  mutate(FinalFC = as.numeric(FinalFC)) %>%
  {
addWorksheet(wbfxn, "add FinalFC") 
writeData(wbfxn,sheet = "add FinalFC", .)

  .}%>%
  # exlude the following columns: plusFC, negFC, and FC
  dplyr::select(-plusFC, -negFC, -FC) %>%
  relocate(FinalFC, .before = 4) %>%
  {
addWorksheet(wbfxn, "exclude plusFC, negFC, intFC") 
writeData(wbfxn, sheet = "exclude plusFC, negFC, intFC", .)


  .}%>%
  #keep genes if the FinalFC is greater than +2 or less than -2
  filter(FinalFC > 2 | FinalFC < -2) %>%
  #keep genes if the log2foldchange is greater than +1 or less than -1
  filter(log2FoldChange > 1 | log2FoldChange < -1) %>%
  {
addWorksheet(wbfxn, "FinalFC > 2, <-2") 
writeData(wbfxn, sheet = "FinalFC > 2, <-2", .)

  .}
    tibble_exp
  file_path <- file.path(getwd(), paste0(file_name, ".xlsx"))
  
worksheetOrder(wbfxn) <- c(1:length(wbfxn$worksheets))
saveWorkbook(wbfxn, file_path, overwrite = TRUE)
  return(list(tibble_exp = tibble_exp, file_name = file_name))
}
```

Taking into consideration the if the values not bee reversed
```{r}
object <- deseq_filteration(tibble =  results_frame, file_name = "test1_2", pval_cutoff = 0.01, rev = FALSE)
```

Taking into consideration if the values have been reversed
```{r}
object2 <- deseq_filteration(tibble =  results_frame, file_name = "test1_2", pval_cutoff = 0.01, rev = TRUE)
```

```{r}
genes <- object2$tibble_exp %>% rownames()
genes
```



Using the Deseq_filteration function then converting the genenames to column for data prep into enrichr function
```{r}
object1 <- deseq_filteration(tibble =  results_frame, file_name = "test1_2", pval_cutoff = 0.01, rev = FALSE)
object1 <- object1$tibble_exp %>% tibble::rownames_to_column("entrezgene_id")
object1
```
Using enrichGO to get results Common exmaple is below
```{r}
key <- enrichGO(
  genes,                    
  "org.Hs.eg.db",           
  ont = "BP",                
  pvalueCutoff = 0.05,      
  pAdjustMethod = "BH"      
)
enrichgo_result <- key@result
```


Get to know you enrichr results and choose the organism and description to extract
use enrichgo unpack to get you results in a readable item
```{r}
enrichgo_result <- key@result
chosen_desc <- "stress response to metal ion"
ensemble_dataset <- "hsapiens_gene_ensembl"
deseq2_table <- object1
enrichgo_unpack  <- enrichgo_unpack(deseq2_table, enrichgo_result, chosen_desc, ensemble_dataset)
```

Display
```{r}
enrichgo_unpack
```



Visual Functions: Must be used after DAVID_Annotation Function

Fxn: Dot Plot

Fxn: Volcano Plot

Required columns: GSEA pvalue, GSEA Term, GSEA Count column

Dot plot post process
```{r}
enrichgo_result <- key@result
gsea_dotplot(enrichgo_result, 0.05, 'Count', 10, "Vignette Testing", file_name = "Vignette_testing")
```
Display by Count or Pvalue

```{r}
enrichgo_result <- key@result
gsea_dotplot(enrichgo_result, 0.05, 'pvalue', 10, "Vignette Testing", file_name = "Vignette_testing")
```

Fxn: Volcano Plot

Required data: your enrichGO results

```{r}
gsea_results <- enrichgo_unpack
```

Volcano Plot visual functions



```{r}
gsea_volplot(gsea_results)
```
Update goal Take in a list of descriptions and have them in the visual plots

As of 4/12/2025 the motiff-finding is still being worked on.
The RNA-Seq and ChIP-Seq Pipeline are almost finished.
Update ChIP-Seq PP has run with the Tet2 real-big data: WT CUT&TAG

Do do list
Run ChIP-Seq PP on Tet2 KO CUT&TAG data

Rework RNA-Seq pipeline

Dock everything together

MOTIF Finding tutorial almost done.
```{r}
```



