```{r}
library(DESeq2)
library(tidyverse)
```

Input should be unormalized data counts

Data Source : <https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE60424>

```{r}
conditions <- c(
  "Healthy Control", "MS pretreatment", "MS posttreatment", "Type 1 Diabetes", "Type 1 Diabetes",
  "Healthy Control", "Healthy Control", "Healthy Control", "Healthy Control", "Healthy Control",
  "Healthy Control", "MS pretreatment", "MS pretreatment", "MS pretreatment", "MS pretreatment",
  "MS pretreatment", "MS posttreatment", "MS posttreatment", "MS posttreatment", "MS posttreatment",
  "MS posttreatment", "Type 1 Diabetes", "Type 1 Diabetes", "Type 1 Diabetes", "Type 1 Diabetes",
  "Type 1 Diabetes", "Type 1 Diabetes", "Type 1 Diabetes", "Type 1 Diabetes", "Type 1 Diabetes",
  "Type 1 Diabetes", "Type 1 Diabetes", "Sepsis", "Sepsis", "Sepsis", "Sepsis", "Sepsis", "Sepsis",
  "Sepsis", "Sepsis", "Sepsis", "Sepsis", "Sepsis", "Sepsis", "Sepsis", "Sepsis", "Sepsis", "Sepsis",
  "Sepsis", "Sepsis", "Sepsis", "Sepsis", "MS pretreatment", "MS pretreatment", "MS pretreatment",
  "MS pretreatment", "MS pretreatment", "MS pretreatment", "MS pretreatment", "MS posttreatment",
  "MS posttreatment", "MS posttreatment", "MS posttreatment", "MS posttreatment", "MS posttreatment",
  "MS posttreatment", "Healthy Control", "Healthy Control", "Healthy Control", "Healthy Control",
  "Healthy Control", "Healthy Control", "Healthy Control", "Healthy Control", "Healthy Control",
  "Healthy Control", "Healthy Control", "Healthy Control", "Healthy Control", "Healthy Control",
  "ALS", "ALS", "ALS", "ALS", "ALS", "ALS", "ALS", "Healthy Control", "Healthy Control",
  "Healthy Control", "Healthy Control", "Healthy Control", "Healthy Control", "Healthy Control",
  "ALS", "ALS", "ALS", "ALS", "ALS", "ALS", "ALS", "Type 1 Diabetes", "Type 1 Diabetes",
  "Type 1 Diabetes", "Type 1 Diabetes", "Type 1 Diabetes", "Type 1 Diabetes", "Type 1 Diabetes",
  "Type 1 Diabetes", "Type 1 Diabetes", "Type 1 Diabetes", "Type 1 Diabetes", "Type 1 Diabetes",
  "Type 1 Diabetes", "MS pretreatment", "MS pretreatment", "MS pretreatment", "MS pretreatment",
  "MS pretreatment", "MS pretreatment", "MS posttreatment", "MS posttreatment", "MS posttreatment",
  "MS posttreatment", "MS posttreatment", "MS posttreatment", "MS posttreatment", "ALS", "ALS",
  "ALS", "ALS", "ALS", "ALS", "ALS"
)
```

```{r}
#load in r
toy_data <- read_delim("~/bioinf-576/toy_data/GSE60424_raw_counts_GRCh38.p13_NCBI.tsv", delim = "\t")
toy_data <- as.data.frame(toy_data)
#Set GeneID as row names 
counts <- column_to_rownames(toy_data, var = "GeneID")

#Create metadata 
metadata <- data.frame(condition = conditions, row.names = colnames(counts))
```

User inputs: countsData = user read in data, colData = metadata, design = conditions of each sample included.

Note: Inputs: countData = user read in data,

```{r}
dds <- DESeqDataSetFromMatrix(
  countData = counts,
  colData = metadata,
  design = ~ condition
)
dds <- DESeq(dds)

normalized_counts <- counts(dds, normalized = TRUE)

results_table <- results(dds)

results_frame <- as.data.frame(results_table)
```

```{r}
results_frame <- as.data.frame(results_table)
results_frame
```

Zeroth Function –\> Chip_seq and RNA_seq Merger (Requirements: same experiment/organism and annotations)

First Function: deseq filtration

step 1, applying pvalue arguments

provisional function data to txt

```{r}
df_to_txt <- function(tibble) {
  
  tibble_name <- deparse(substitute(tibble))
  
  file_name <- paste(getwd(),"/",tibble_name,".txt")
  
  file_mod <- gsub("\\s", "", file_name)
  #\\s is a regular expression pattern that matches any white space character
  #gsub replaces all occurance of this pattern with an empty string
  save_fxn <- write_delim(tibble, file = file_mod, delim = "\t", col_names = FALSE)
  
  return(save_fxn)
}
  #there has to be a return value
```

```{r}
deseq_filteration <- function(tibble, 
                              file_name, 
                              pval_cutoff, 
                              rev = FALSE){
#added argument: pval_cutoff, allows user to set their pvalue cutoff
# This function takes into consideration that the Deseq2 Results where flipped
# I'll have to make a conditional on that front
#
library(tidyverse)
  
  if(!require(openxlsx)) install.packages("openxlsx", dependencies=TRUE)
  
  if(!require(dplyr)) install.packages("dplyr", dependencies=TRUE)
  
library(dplyr)
library(openxlsx)
  
wbfxn <- createWorkbook()

  addWorksheet(wbfxn, paste("Initial Data", nrow(tibble)))
  writeData(wbfxn, sheet = paste("Initial Data", nrow(tibble)) , tibble)

tibble_exp <- tibble %>%
# filter in value <=0.05
  
#In the past I have encountered Deseq sets wherelog2FCs where the results are flipped
#This conditional takes that into consideration
  {
    if (rev) {
      . <- mutate(., INUN_log2FC = log2FoldChange * -1)
      . <- relocate(., INUN_log2FC, .after = 2)
      addWorksheet(wbfxn, paste("Corrected Log2FC", nrow(.)))
      writeData(wbfxn, sheet = paste("Corrected Log2FC", nrow(.)), .)
    } else {
      . <- mutate(., log2FoldChange = log2FoldChange)
    }
    
  .}%>%
  filter(pvalue < pval_cutoff) %>%
  {
    addWorksheet(wbfxn, paste("pvalue <=", pval_cutoff, nrow(.)))
    writeData(wbfxn, sheet = paste("pvalue <=", pval_cutoff, nrow(.)), .)
    
  .}%>%
  
  {
    if ("INUN_log2FC" %in% colnames(.)) {
      mutate(., plusFC = 2^INUN_log2FC)
    } else {
      mutate(., plusFC = 2^log2FoldChange)
    }
  } %>%
  
  relocate(plusFC, .before = 4) %>%
  {
    
addWorksheet(wbfxn, "add plusFC") 
writeData(wbfxn,sheet = "add plusFC", .) 

  .}%>%
  
  arrange(desc(plusFC)) %>%
  {
addWorksheet(wbfxn, "Sort by plusFC desc.") 
writeData(wbfxn, sheet = "Sort by plusFC desc.", .)

  .}%>%
  
#add a column called negFC: if a value in plusFC is less than 0.05 then convert it to
#negFC (-1/FC) else leave the value alone by multiplying it by 1
  mutate(negFC = if_else(plusFC < 0.5, -1/plusFC, plusFC * 1)) %>%
  relocate(negFC, .before = 5) %>%
  {
addWorksheet(wbfxn, "add negFC") 
writeData(wbfxn, sheet = "add negFC", .)
  .}%>%
  
#In a new column called FC, 
#If the values in plusFC and negativeFC match (plusFC == negFC) then copy the values that do match into the new column and convert them to the character type 
#Else: paste both values from negFC and plusFC into the same point and seperate them by a "/" character.
  mutate(FC = if_else(plusFC == negFC, as.character(plusFC), paste(plusFC, negFC, sep = "/"))) %>%
  #In the new column FinalFC, If an entry has a slash charcter in the column FC, then get rid of everycharacter before the slash in each entry including the slash.
  mutate(FinalFC = gsub(".*?/","", FC)) %>%
  #convert the column FinalFC into a numeric type variable
  #The goal of the FinalFC column is only keep the strongest directional fold change
  mutate(FinalFC = as.numeric(FinalFC)) %>%
  {
addWorksheet(wbfxn, "add FinalFC") 
writeData(wbfxn,sheet = "add FinalFC", .)

  .}%>%
  # exlude the following columns: plusFC, negFC, and FC
  dplyr::select(-plusFC, -negFC, -FC) %>%
  relocate(FinalFC, .before = 4) %>%
  {
addWorksheet(wbfxn, "exclude plusFC, negFC, intFC") 
writeData(wbfxn, sheet = "exclude plusFC, negFC, intFC", .)


  .}%>%
  #keep genes if the FinalFC is greater than +2 or less than -2
  filter(FinalFC > 2 | FinalFC < -2) %>%
  #keep genes if the log2foldchange is greater than +1 or less than -1
  filter(log2FoldChange > 1 | log2FoldChange < -1) %>%
  {
addWorksheet(wbfxn, "FinalFC > 2, <-2") 
writeData(wbfxn, sheet = "FinalFC > 2, <-2", .)

  .}
    tibble_exp
  file_path <- file.path(getwd(), paste0(file_name, ".xlsx"))
  
worksheetOrder(wbfxn) <- c(1:length(wbfxn$worksheets))
saveWorkbook(wbfxn, file_path, overwrite = TRUE)
  return(list(tibble_exp = tibble_exp, file_name = file_name))
}
```

```{r}
object <- deseq_filteration(tibble =  results_frame, file_name = "test1_2", pval_cutoff = 0.01, rev = FALSE)
```

```{r}
object2 <- deseq_filteration(tibble =  results_frame, file_name = "test1_2", pval_cutoff = 0.01, rev = TRUE)
```

```{r}
object2$tibble_exp
```

DAVID Annotation Function FUN FUN FUN

```{r}

```

Visual Functions: Must be used after DAVID_Annotation Function

Fxn: Dot Plot

Fxn: Volcano Plot

Required columns: GSEA pvalue, GSEA Term, GSEA Count column

```{r}
#arguments
#Required columns: GSEA pvalue, GSEA Term, GSEA Count column
#organize by Count or pvalue GSEA results (org = 'Count', org = 'pvalue')
#num_disp
#GSEA pvalue filter
#graph title (title)
#file_name write out
#output visual dotplot of GSEA enrichment
#input post GSEA

gsea_dotplot <- function(gsea_results, gspval_cutoff, org_by, num_disp, graph_title, file_name){
  if(org_by %in% 'Count'){
    #inside if
    dotplot1 <- gsea_resuts%>%
                arrange(desc(Count))%>%
                head(num_disp)%>%
                mutate(as.factor(Term))%>%
                ggplot() +
                aes(y = Count, x = fct_reorder(Term, Count), fill = pvalue) +
                scale_fill_gradient(low = "red", high = "blue") +
                geom_col(color = "black") +
                labs(title = graph_title, x = "Enriched Term", y = "Gene Count") +
                coord_flip() +
                theme(text = element_text(face = "bold") , plot.title = element_text(hjust = 1))
    
  }
  if(org_by %in% 'pvalue'){
    #inside if
    dotplot2 <- gsea_resuts %>%
                filter(pvalue < gspval_cutoff) %>%
                head(num_disp) %>%
                mutate(as.factor(Term)) %>%
                mutate(pvalue = -1 * log(pvalue)) %>%
                ggplot() +
                aes(x = pvalue, y = fct_reorder(Term, pvalue), fill = Count) +
                scale_fill_gradient(low = "red", high = "blue") +
                geom_col(color = "black") +
                labs(title = graph_title, x = "-log10(P)", y = "Enriched Term") +
                theme(text = element_text(face = "bold"), plot.title = element_text(hjust = 1))
  
              
  }
  return(list(dotplot, ggsave(file_name, device = "png", width = 8, height = 6, units = "in")))
}
```

Fxn: Volcano Plot

Required data:

```{r}
#'arguments: gsea_results, 
#'gsea_result desc: should have following columns: a gene_id, gsea_pvalue, gsea_term,gsea_count
```

```{r}

#setting the color
gsea_results$color <- ifelse(gsea_results$FinalFC > 2, "Upregulated",  "Downregulated")

#arrange by padj, may give that option to the user
#the idea is that we are using DAVID so that means we some strict
#It looks like this function is under the assumption that annotation
gsea_results <- gsea_results %>%
  arrange(padj) %>%
  mutate(delabel = ifelse(row_number() <= 10, as.character(gene_id_v1.2), ""))


ggplot(gsea_results) +
  aes(x = FinalFC, y = -log10(pvalue), color = color, label = delabel) +
  
  
  geom_vline(xintercept = c(-0.6, 0.6), col = "gray", linetype = 'dashed') +
  
  geom_hline(yintercept = -log10(0.05), col = "gray", linetype = 'dashed') + 
  
  theme_set(theme_classic(base_size = 12) +
              
  theme(
  axis.title.y = element_text(margin = margin(0,20,0,0), size = rel(1.1), color = 'black'),
  
  axis.title.x = element_text(hjust = 0.5, margin = margin(20,0,0,0), size = rel(1.1), color = 'black'),

  plot.title = element_text(hjust = 0.5)
            )) +
  
  geom_point(size = 2) +
  
  scale_color_manual(values = c("Upregulated" = "red", "Downregulated" = "blue")) +
  
  geom_text(nudge_y = 0.5, check_overlap = TRUE) +
  
  geom_text_repel(max.overlaps = Inf) +
  coord_cartesian(xlim = c(-30, 30)) +
  scale_x_continuous(breaks = seq(-30, 30, 5))
)
  
```

```{r}
gsea_volplot <- function(gsea_results){
       #setting the color
    gsea_results$color <- ifelse(gsea_results$FinalFC > 2, "Upregulated",  "Downregulated")
    
    #arrange by padj, may give that option to the user
    #the idea is that we are using DAVID so that means we some strict
    #It looks like this function is under the assumption that annotation
    gsea_results <- gsea_results %>%
      arrange(padj) %>%
      mutate(delabel = ifelse(row_number() <= 10, as.character(gene_id_v1.2), ""))
    
    
    ggplot(gsea_results) +
      aes(x = FinalFC, y = -log10(pvalue), color = color, label = delabel) +
      
      
      geom_vline(xintercept = c(-0.6, 0.6), col = "gray", linetype = 'dashed') +
      
      geom_hline(yintercept = -log10(0.05), col = "gray", linetype = 'dashed') + 
      
      theme_set(theme_classic(base_size = 12) +
                  
      theme(
      axis.title.y = element_text(margin = margin(0,20,0,0), size = rel(1.1), color = 'black'),
      
      axis.title.x = element_text(hjust = 0.5, margin = margin(20,0,0,0), size = rel(1.1), color = 'black'),
    
      plot.title = element_text(hjust = 0.5)
                )) +
      
      geom_point(size = 2) +
      
      scale_color_manual(values = c("Upregulated" = "red", "Downregulated" = "blue")) +
      
      geom_text(nudge_y = 0.5, check_overlap = TRUE) +
      
      geom_text_repel(max.overlaps = Inf) +
      coord_cartesian(xlim = c(-30, 30)) +
      scale_x_continuous(breaks = seq(-30, 30, 5)) 
  return(gsea_results)
}
```
