```{r}
library(DESeq2)
library(tidyverse)
library(clusterProfiler)
library(org.Hs.eg.db)
library(biomaRt)
```

```{r}
#this function needs to be generalized. How? Need to take a page out of the clusterprofiler
mmu_entrezid_to_genename <- function(entrez_vector) {
  #requires both of these packages to run
  #require("biomaRt")
  #require("org.Mmu.eg.db")
  #connect to the Ensembl database for Mmu
  ensembl = useMart("ensembl", dataset = "mmusculus_gene_ensembl")
  #getting the corresponding Entrez Gene IDs
  entrezid_genename_df <- getBM(attributes = c("entrezgene_id", "external_gene_name"), filters = "entrezgene_id", values = entrez_vector, mart = ensembl)
  #return both the df and the vector
  return(entrezid_genename_df)
}
```

```{r}
##' @param
##' @param OrgDb OrgDb
##' @import biomaRt
# common example datasets: "hsapiens_gene_ensembl", "mmusculus_gene_ensembl"
#connects to the ensembledb, grabs and returns
#arguments: an entrez id vector, and the name of ensembldataset
#It would be wise to load in your chosen OrgDb set before running this code


entrezid_to_genename <-function(entrez_vector,
                                ensembl_dataset = "hsapiens_gene_ensembl"){
                          #connect to the Ensembl database for Mmu
                          ensembl = useMart("ensembl", dataset = ensembl_dataset)
                          #getting the corresponding Entrez Gene IDs
                          entrezid_genename_df <- getBM(attributes = c("entrezgene_id", "external_gene_name"), filters = "entrezgene_id", values = entrez_vector, mart = ensembl)
    return(entrezid_genename_df)
}
```

Input should be unormalized data counts

Data Source : <https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE60424>

Importing Meta Data cats()

```{r}
conditions <- c(
  "Healthy Control", "MS pretreatment", "MS posttreatment", "Type 1 Diabetes", "Type 1 Diabetes",
  "Healthy Control", "Healthy Control", "Healthy Control", "Healthy Control", "Healthy Control",
  "Healthy Control", "MS pretreatment", "MS pretreatment", "MS pretreatment", "MS pretreatment",
  "MS pretreatment", "MS posttreatment", "MS posttreatment", "MS posttreatment", "MS posttreatment",
  "MS posttreatment", "Type 1 Diabetes", "Type 1 Diabetes", "Type 1 Diabetes", "Type 1 Diabetes",
  "Type 1 Diabetes", "Type 1 Diabetes", "Type 1 Diabetes", "Type 1 Diabetes", "Type 1 Diabetes",
  "Type 1 Diabetes", "Type 1 Diabetes", "Sepsis", "Sepsis", "Sepsis", "Sepsis", "Sepsis", "Sepsis",
  "Sepsis", "Sepsis", "Sepsis", "Sepsis", "Sepsis", "Sepsis", "Sepsis", "Sepsis", "Sepsis", "Sepsis",
  "Sepsis", "Sepsis", "Sepsis", "Sepsis", "MS pretreatment", "MS pretreatment", "MS pretreatment",
  "MS pretreatment", "MS pretreatment", "MS pretreatment", "MS pretreatment", "MS posttreatment",
  "MS posttreatment", "MS posttreatment", "MS posttreatment", "MS posttreatment", "MS posttreatment",
  "MS posttreatment", "Healthy Control", "Healthy Control", "Healthy Control", "Healthy Control",
  "Healthy Control", "Healthy Control", "Healthy Control", "Healthy Control", "Healthy Control",
  "Healthy Control", "Healthy Control", "Healthy Control", "Healthy Control", "Healthy Control",
  "ALS", "ALS", "ALS", "ALS", "ALS", "ALS", "ALS", "Healthy Control", "Healthy Control",
  "Healthy Control", "Healthy Control", "Healthy Control", "Healthy Control", "Healthy Control",
  "ALS", "ALS", "ALS", "ALS", "ALS", "ALS", "ALS", "Type 1 Diabetes", "Type 1 Diabetes",
  "Type 1 Diabetes", "Type 1 Diabetes", "Type 1 Diabetes", "Type 1 Diabetes", "Type 1 Diabetes",
  "Type 1 Diabetes", "Type 1 Diabetes", "Type 1 Diabetes", "Type 1 Diabetes", "Type 1 Diabetes",
  "Type 1 Diabetes", "MS pretreatment", "MS pretreatment", "MS pretreatment", "MS pretreatment",
  "MS pretreatment", "MS pretreatment", "MS posttreatment", "MS posttreatment", "MS posttreatment",
  "MS posttreatment", "MS posttreatment", "MS posttreatment", "MS posttreatment", "ALS", "ALS",
  "ALS", "ALS", "ALS", "ALS", "ALS"
)
```

Importing Raw Counts Data

User must assemble the following pieces of data before using the functions in this package:

1: They must have pre-processed RNa-Seq in a raw un-normalized count matrix and ChIP-Seq data in the Annotated Peak Call Form

2: They must have meta data which is equal in length to the sample name vector (Use will probably have to search.

```{r}
#load in r
toy_data <- read_delim("~/bioinf-576/toy_data/GSE60424_raw_counts_GRCh38.p13_NCBI.tsv", delim = "\t")
toy_data <- as.data.frame(toy_data)
#Set GeneID as row names 
counts <- column_to_rownames(toy_data, var = "GeneID")

#Create metadata 
metadata <- data.frame(condition = conditions, row.names = colnames(counts))
```

User inputs: countsData = user read in data, colData = metadata, design = conditions of each sample included.

Note: Inputs: countData = user read in data,

```{r}
dds <- DESeqDataSetFromMatrix(
  countData = counts,
  colData = metadata,
  design = ~ condition
)
dds <- DESeq(dds)

normalized_counts <- counts(dds, normalized = TRUE)

results_table <- results(dds)

results_frame <- as.data.frame(results_table)
```

```{r}
results_frame <- as.data.frame(results_table)
results_frame
```

Zeroth Function Chip_seq and RNA_seq Merger (Requirements: same experiment/organism and annotations)

First Function: deseq filtration

step 1, applying pvalue arguments

provisional function data to txt

```{r}
deseq_filteration <- function(tibble, 
                              file_name, 
                              pval_cutoff, 
                              rev = FALSE){
#added argument: pval_cutoff, allows user to set their pvalue cutoff
# This function takes into consideration that the Deseq2 Results where flipped
# I'll have to make a conditional on that front
#
library(tidyverse)
  
  if(!require(openxlsx)) install.packages("openxlsx", dependencies=TRUE)
  
  if(!require(dplyr)) install.packages("dplyr", dependencies=TRUE)
  
library(dplyr)
library(openxlsx)
  
wbfxn <- createWorkbook()

  addWorksheet(wbfxn, paste("Initial Data", nrow(tibble)))
  writeData(wbfxn, sheet = paste("Initial Data", nrow(tibble)) , tibble)

tibble_exp <- tibble %>%
# filter in value <=0.05
  
#In the past I have encountered Deseq sets wherelog2FCs where the results are flipped
#This conditional takes that into consideration
  {
    if (rev) {
      . <- mutate(., INUN_log2FC = log2FoldChange * -1)
      . <- relocate(., INUN_log2FC, .after = 2)
      addWorksheet(wbfxn, paste("Corrected Log2FC", nrow(.)))
      writeData(wbfxn, sheet = paste("Corrected Log2FC", nrow(.)), .)
    } else {
      . <- mutate(., log2FoldChange = log2FoldChange)
    }
    
  .}%>%
  filter(pvalue < pval_cutoff) %>%
  {
    addWorksheet(wbfxn, paste("pvalue <=", pval_cutoff, nrow(.)))
    writeData(wbfxn, sheet = paste("pvalue <=", pval_cutoff, nrow(.)), .)
    
  .}%>%
  
  {
    if ("INUN_log2FC" %in% colnames(.)) {
      mutate(., plusFC = 2^INUN_log2FC)
    } else {
      mutate(., plusFC = 2^log2FoldChange)
    }
  } %>%
  
  relocate(plusFC, .before = 4) %>%
  {
    
addWorksheet(wbfxn, "add plusFC") 
writeData(wbfxn,sheet = "add plusFC", .) 

  .}%>%
  
  arrange(desc(plusFC)) %>%
  {
addWorksheet(wbfxn, "Sort by plusFC desc.") 
writeData(wbfxn, sheet = "Sort by plusFC desc.", .)

  .}%>%
  
#add a column called negFC: if a value in plusFC is less than 0.05 then convert it to
#negFC (-1/FC) else leave the value alone by multiplying it by 1
  mutate(negFC = if_else(plusFC < 0.5, -1/plusFC, plusFC * 1)) %>%
  relocate(negFC, .before = 5) %>%
  {
addWorksheet(wbfxn, "add negFC") 
writeData(wbfxn, sheet = "add negFC", .)
  .}%>%
  
#In a new column called FC, 
#If the values in plusFC and negativeFC match (plusFC == negFC) then copy the values that do match into the new column and convert them to the character type 
#Else: paste both values from negFC and plusFC into the same point and seperate them by a "/" character.
  mutate(FC = if_else(plusFC == negFC, as.character(plusFC), paste(plusFC, negFC, sep = "/"))) %>%
  #In the new column FinalFC, If an entry has a slash charcter in the column FC, then get rid of everycharacter before the slash in each entry including the slash.
  mutate(FinalFC = gsub(".*?/","", FC)) %>%
  #convert the column FinalFC into a numeric type variable
  #The goal of the FinalFC column is only keep the strongest directional fold change
  mutate(FinalFC = as.numeric(FinalFC)) %>%
  {
addWorksheet(wbfxn, "add FinalFC") 
writeData(wbfxn,sheet = "add FinalFC", .)

  .}%>%
  # exlude the following columns: plusFC, negFC, and FC
  dplyr::select(-plusFC, -negFC, -FC) %>%
  relocate(FinalFC, .before = 4) %>%
  {
addWorksheet(wbfxn, "exclude plusFC, negFC, intFC") 
writeData(wbfxn, sheet = "exclude plusFC, negFC, intFC", .)


  .}%>%
  #keep genes if the FinalFC is greater than +2 or less than -2
  filter(FinalFC > 2 | FinalFC < -2) %>%
  #keep genes if the log2foldchange is greater than +1 or less than -1
  filter(log2FoldChange > 1 | log2FoldChange < -1) %>%
  {
addWorksheet(wbfxn, "FinalFC > 2, <-2") 
writeData(wbfxn, sheet = "FinalFC > 2, <-2", .)

  .}
    tibble_exp
  file_path <- file.path(getwd(), paste0(file_name, ".xlsx"))
  
worksheetOrder(wbfxn) <- c(1:length(wbfxn$worksheets))
saveWorkbook(wbfxn, file_path, overwrite = TRUE)
  return(list(tibble_exp = tibble_exp, file_name = file_name))
}
```

```{r}
object <- deseq_filteration(tibble =  results_frame, file_name = "test1_2", pval_cutoff = 0.01, rev = FALSE)
```

```{r}
object2 <- deseq_filteration(tibble =  results_frame, file_name = "test1_2", pval_cutoff = 0.01, rev = TRUE)
```

```{r}
genes <- object2$tibble_exp %>% rownames()
genes
```

```{r}
object1 <- deseq_filteration(tibble =  results_frame, file_name = "test1_2", pval_cutoff = 0.01, rev = FALSE)
object1 <- object1$tibble_exp %>% tibble::rownames_to_column("entrezgene_id")
object1
```

We can't find the DAVID annotatiion stuff I have to use clusterProfiler

```{r}
genes <- object1$entrezgene_id
```

```{r}
key <- enrichGO(
  genes,                    
  "org.Hs.eg.db",           
  ont = "BP",                
  pvalueCutoff = 0.05,      
  pAdjustMethod = "BH"      
)
```

Unpacking the Gene IDs

```{r}
key@result
```

```{r}

##' @import biomaRt
# common example datasets: "hsapiens_gene_ensembl", "mmusculus_gene_ensembl"
#connects to the ensembledb, grabs and returns
#arguments: an entrez id vector, and the name of ensembldataset
#It would be wise to load in your chosen OrgDb set before running this code


entrezid_to_genename <-function(entrez_vector,
                                ensembl_dataset = "hsapiens_gene_ensembl"){
                          #connect to the Ensembl database for Mmu
                          ensembl = useMart("ensembl", dataset = ensembl_dataset)
                          #getting the corresponding Entrez Gene IDs
                          entrezid_genename_df <- getBM(attributes = c("entrezgene_id", "external_gene_name"), filters = "entrezgene_id", values = entrez_vector, mart = ensembl)
    return(entrezid_genename_df)
}
```

enrichgo_unpack

```{r}
deseq2_table <- object1
enrichgo_result <- key@result
chosen_desc <- "cellular response to copper ion"
ensemble_dataset <- "hsapiens_gene_ensembl"

enrichgo_unpack <- function(deseq2_table, enrichgo_result, chosen_desc, ensembl_dataset){
    ###Extracting a specific gene into a cluster
    
    cluster <- tibble(key@result) %>%
      filter(str_detect(Description, chosen_desc)) %>%
      mutate(Description = as.character(Description))
    cluster
    
    ###creates a helper function to unlist and strsplits the enriched Gene IDs per term
    
    convert_to_integer <- function(str) {
      bruh <- as.integer(unlist(strsplit(str, "/")))
      return(bruh)
    }
    
    #apply this function to the geneID column
    extract_result <- lapply(cluster$geneID, convert_to_integer)
    #Next check the length of the new vector to see if it matches the Count column in the GSEA DF.
    length(extract_result[[1]])
    
    #create an empty vector
    path_vector <- c()
    #for each index in the first column of numbers
    for (i in 1) {
      #replicate the name of the Description for the length of the GO term in column 1
      path_vector <- c(path_vector, rep(cluster$Description[i], length(cluster[[i]])))
    }
    
    # make new tibble that pairs the name of the Pathway from the GOTerm and enriched Entrez ID from the "post-processed" GeneID column
    pathway_tibble <- tibble(gene_desc = path_vector, entrezgene_id = unlist(extract_result))
    pathway_tibble
    
    pathway_genenames <- entrezid_to_genename(pathway_tibble$entrezgene_id, ensembl_dataset = ensemble_dataset) 
    
    
    
    final_path_tibble <- inner_join(pathway_tibble, pathway_genenames, by = "entrezgene_id")
    final_path_tibble <- final_path_tibble %>% mutate(entrezgene_id = as.character(entrezgene_id))
    final_path_tibble
    
    final_path_fc_tibble <- inner_join(final_path_tibble, deseq2_table, by = "entrezgene_id")
    final_path_fc_tibble
    return(final_path_fc_tibble)
}
```

```{r}
enrichgo_result <- key@result
chosen_desc <- "stress response to metal ion"
ensemble_dataset <- "hsapiens_gene_ensembl"
deseq2_table <- object1
enrichgo_unpack  <- enrichgo_unpack(deseq2_table, enrichgo_result, chosen_desc, ensemble_dataset)
```

Visual Functions: Must be used after DAVID_Annotation Function

Fxn: Dot Plot

Fxn: Volcano Plot

Required columns: GSEA pvalue, GSEA Term, GSEA Count column

```{r}
#arguments
#Required columns: GSEA pvalue, GSEA Term, GSEA Count column
#organize by Count or pvalue GSEA results (org = 'Count', org = 'pvalue')
#num_disp
#GSEA pvalue filter
#graph title (title)
#file_name write out
#output visual dotplot of GSEA enrichment
#input post GSEA

gsea_dotplot <- function(gsea_results, gspval_cutoff, org_by, num_disp, graph_title, file_name){
  if(org_by %in% 'Count'){
    #inside if
    dotplot <- gsea_results%>%
                arrange(desc(Count))%>%
                head(num_disp)%>%
                mutate(as.factor(Description))%>%
                ggplot() +
                aes(y = Count, x = fct_reorder(Description, Count)) +
                scale_colour_gradient(low = "red", high = "blue") +
                geom_point(aes(size = GeneRatio, color = pvalue)) +
                labs(title = graph_title, x = "Enriched Term", y = "Gene Count") +
                coord_flip() +
                theme(text = element_text(face = "bold") , plot.title = element_text(hjust = 1))
    
  }
  if(org_by %in% 'pvalue'){
    #inside if
    dotplot <- gsea_results %>%
                filter(pvalue < gspval_cutoff) %>%
                head(num_disp) %>%
                mutate(as.factor(Description)) %>%
                mutate(pvalue = -1 * log(pvalue)) %>%
                ggplot() +
                aes(x = pvalue, y = fct_reorder(Description, pvalue))+
                scale_colour_gradient(low = "red", high = "blue") +
                geom_point(aes(size = GeneRatio, color = Count)) +
                labs(title = graph_title, x = "-log10(P)", y = "Enriched Term") +
                theme(text = element_text(face = "bold"), plot.title = element_text(hjust = 1))
  
              
  }
  return(list(dotplot, ggsave(file_name, device = "png", width = 8, height = 6, units = "in")))
}
```

```{r}
enrichgo_result <- key@result
gsea_dotplot(enrichgo_result, 0.05, 'pvalue', 10, "graph_title", "file_name")
```

```{r}
enrichgo_result <- key@result
enrichgo_result
```

Fxn: Volcano Plot

Required data:

```{r}
#'arguments: gsea_results, 
#'gsea_result desc: should have following columns: a gene_id, gsea_pvalue, gsea_term,gsea_count
```

```{r}
gsea_results <- enrichgo_unpack
```

```{r}
gsea_volplot <- function(gsea_results){
       #setting the color
      
    gsea_results$color <- ifelse(gsea_results$log2FoldChange > 0, "Upregulated",  "Downregulated")
    
    #arrange by padj, may give that option to the user
    #the idea is that we are using DAVID so that means we some strict
    #It looks like this function is under the assumption that annotation
    gsea_results <- gsea_results %>%
      arrange(padj) %>%
      mutate(delabel = ifelse(row_number() <= 10, as.character(external_gene_name), ""))
    
    
    volplot <- ggplot(gsea_results) +
      aes(x = log2FoldChange, y = -log10(pvalue), color = color, label = delabel) +
      
      
      geom_vline(xintercept = c(-0.6, 0.6), col = "gray", linetype = 'dashed') +
      
      geom_hline(yintercept = -log10(0.05), col = "gray", linetype = 'dashed') + 
      
      theme_set(theme_classic(base_size = 12) +
                  
      theme(
      axis.title.y = element_text(margin = margin(0,20,0,0), size = rel(1.1), color = 'black'),
      
      axis.title.x = element_text(hjust = 0.5, margin = margin(20,0,0,0), size = rel(1.1), color = 'black'),
    
      plot.title = element_text(hjust = 0.5)
                )) +
      
      geom_point(size = 2) +
      
      scale_color_manual(values = c("Upregulated" = "red", "Downregulated" = "blue")) +
      
      geom_text(nudge_y = 0.5, check_overlap = TRUE) +
      
      geom_text_repel(max.overlaps = Inf) +
      coord_cartesian(xlim = c(-30, 30)) +
      scale_x_continuous(breaks = seq(-30, 30, 5))
    
  return(list(volplot, gsea_results))
}
```

```{r}
gsea_volplot(gsea_results)
```
